Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-08-31 jso.qmd
adding sec6 to intro paragraph
typo
revised abstract

  The projection pursuit (PP) guided tour interactively optimises a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. The optimisation in PP can be non-trivial, involving non-smooth functions and optima with a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated along with JSO hyper-parameters to determine their effects on JSO success rate. Our numerical results confirm the positive impact of these metrics on the JSO success rate, with squintability being the most significant. The JSO algorithm has been implemented in the `tourr` package and functions to calculate smoothness and squintability are available in the `ferrn` package. 

@@

The paper is structured as follows. @sec-background introduces the
background of PP guided tour and reviews existing optimisers and index
functions in the literature. @sec-theory details JSO and introduces
metrics that measure different properties of PP indexes, smoothness and
squintability. @sec-sim-deets details two simulation experiments to
assess JSO's performance: one comparing JSO's performance improvements
relative to an existing optimiser, Creeping Random Search (CRS), and the
other studying the impact of different PP index properties on
optimisation performance. @sec-sim-res presents the results.
@sec-conclusion summarises the work and provides suggestions for future
directions.

@@

The JSO performance is compared with an existing optimiser, Creeping
Random Search (CRS) [@RJ-2021-105; @laa_using_2020] used in the PP
guided tour to explore JSO's behaviour under different hyper-parameter
and data dimension combinations. The second simulation studies the
effect of index properties (smoothness and squintability), along with
JSO hyper-parameters, and data dimension, on the success rate of the JSO
performance. This section describes the simulation details, with the
results deferred to Section @sec-sim-res.

>>>>

  The projection pursuit (PP) guided tour interactively optimises a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optima has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. There are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour.  The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 

@@

The paper is structured as follows. @sec-background introduces the
background of PP guided tour and reviews existing optimisers and index
functions in the literature. @sec-theory details JSO and introduces
metrics that measure different properties of PP indexes, smoothness and
squintability. @sec-sim-deets details two simulation experiments to
assess JSO's performance: one comparing JSO's performance improvements
relative to an existing optimiser, Creeping Random Search (CRS), and the
other studying the impact of different PP index properties on
optimisation performance. @sec-sim-res presents the results. @sec-discussion discusses implementation details and insights for practitioners. 
@sec-conclusion summarises the work and provides suggestions for future
directions.

@@

The JSO performance is compared with an existing optimiser, Creeping
Random Search (CRS) [@RJ-2021-105; @laa_using_2020] used in the PP
guided tour to explore JSO's behaviour under different hyper-parameter
and data dimension combinations. The second simulation studies the
effect of index properties (smoothness and squintability), along with
JSO hyper-parameters, and data dimension, on the success rate of the JSO
performance. This section describes the simulation details, with the
results deferred to @sec-sim-res.
